Collaborative Visual Analysis on the Web with RCloud

### 2015-06-23 discussion.tex
reduced text on caching and security

\paragraph*{Caching}
An important consideration is how and where to introduce caching
in RCloud's distributed computation model. Caching can dramatically
improve performance in a way that is otherwise transparent to
application program semantics \cite{Callahan:2006:VVM, Guo:2010:TPI}.
Though caching is usually implicit, in some environments, such as
VisTrails, programmers may also have some explicit control over cache
management. This may be desirable to ensure the repeatability of
computations that rely on volatile or unreliable data sources.
For example, the stock ticker use case is one for which the data
might temporarily become unavailable, so caching could improve
reliability and consistency of results. Alternatively, in situations
involving relatively expensive computation (for example, analysis of large
text corpora, clustering multivariate time series or deep learning algorithms)
caching may be essential to adequate performance. Currently we would
program this in RCloud by explicitly saving an analysis in a persistent database,
but it seems better to implement this capability in a general purpose associative
cache, instead of application-specific libraries. We envision cache management
being implemented in a new middle layer to be added in the future.

\paragraph*{Security}
Security is an important concern in an environment for
collaboration and data publishing. To provide some
protection for RCloud workbooks in an organization's intranet,
RCloud uses an object capability model
\cite{Miller:2006:RCT}
recently added to the Rserve protocol \cite{Urbanek:2003:AFW}
that prevents unauthenticated clients from making
unauthorized calls to the RCloud runtime environment.

Our back-end environment assumes a high degree of trust between users.
Access control for information security is delegated to the host
operating system and web server. In practice, most operating
systems and web servers rely either on coarse permission models,
which tend to be ineffective, or on detailed access control
lists, which tend to be cumbersome.
Information security in RCloud should be improved. More
sophisticated approaches, such as information flow analysis,
and query languages that ensure differential privacy, are
active research topics, and many problems remain
\cite{Moore:2011:SAF}.
% We are hopeful that having a richer
% environment for collaboration and information sharing will encourage
% new approaches to information security in visual analytics.
%\gordon{This sounds like we think we are helping by pushing the
%system to the breaking point, is that the idea?}

\paragraph*{Exploration is hard}
% R is a popular language for exploratory data analysis, and we built RCloud
% as a way for data scientists to seamlessly transition from EDA to sharing
% and presentation.
Users reported that RCloud's combined notebook and command line interface
interferes with the familiar use of R for exploration.
In current R practice, the command prompt is used to try things out.
It is very fast and there is often no commitment: nothing gets saved
and there is nothing to clean up.
In contrast, in RCloud's notebook interface, the extra cells need
to be deleted later, and there seems to be a mental burden
that \emph{scratch work is getting saved}.

One thing to try is to make deletion easier, for example with a
shortcut key that removes the last cell, or a way to keep a few cells
and cull the rest. A far more ambitious solution, suggested by one of our
interviewees, is to implement auditing of data analyses as it existed in S
\cite{Becker:1988:Auditing}, so that a result could be selected and code
which did not lead to it could be discarded automatically.

We got here from an insistence on reproducibility: it is a core principle
of RCloud that no command should be run without being saved, even if it
gets deleted later. This is why we did not implement a simple command line
as many users request. We could imagine making an exception for commands
that have no side-effects, but in the general case this is no easier than
auditing because R is not a purely functional language.

The interface is also perceived to be slower, both for computer and for
human. First, there is higher latency when commands run, because
RCloud saves a cell to a repository before executing it. It
may be possible to safely write to the repository behind execution of the
same code. Second, and perhaps more detrimental, the interface requires
switching between the keyboard and mouse, which can be cumbersome.
%\gordon{we don't need to back that up, right? i hate using passive voice.}
Shortcut keys can defined as a workaround, but it would be better
to reduce the need for them.

Other complaints mostly focus on layout, and can probably be addressed by
offering better customization, and designs such as tabbed views.

\paragraph*{Discovery is hard} 
With hundreds of users and thousands of notebooks, the notebook tree
becomes unwieldy. The search function does not protect searchers from
obsolete and erroneous notebooks, which by default are retained
forever.  It helps with finding notebooks, but not with finding other
types of metadata. In other words, \emph{textual search} is easy, but
\emph{relevance search} is what we need. Allowing users to \emph{tag}
notebooks is a natural idea, but it relies on users making a conscious
effort, This goes somewhat against our ideal of not asking the user to
do work for the system. Rather, we believe a better alternative is to
have the system work for the user, relying on passive metrics, such as
how often a notebook has been run.  More generally, infrastructure for
automatic \emph{curation} of the software artifacts generated by teams
seems to be an important avenue for future work.

>>>>

\paragraph*{Caching}
An important consideration is how and where to introduce caching in
RCloud's distributed computation model. Caching can dramatically
improve performance in a way that is otherwise transparent to
application program semantics \cite{Callahan:2006:VVM, Guo:2010:TPI}.
Currently, RCloud offers no support for implicit caching, but since it
supports the entirety of the R language, it's possible to incorporate
caching in R libraries.  Still, since the reuse of intermediate
results is a recurring problem in data exploration environments, it
seems better to implement this capability in a general purpose
associative cache. We envision cache management being implemented in a
new middle layer to be added in the future.

\paragraph*{Security}
Because RCloud allows arbitrary R code to be run
\emph{by design}, it's not straightforward to protect deployments
from running unauthorized arbitrary code. RCloud uses an object capability model
\cite{Miller:2006:RCT} recently added to the Rserve
protocol~\cite{Urbanek:2003:AFW}. Object capabilities are opaque handles
to remote procedure calls; this way, web browsers never directly instruct the
RCloud backend to execute arbitrary code, preventing unauthenticated clients
from making unauthorized calls to the RCloud runtime environment.

Our back-end environment, on the other hand, assumes a high degree of trust between users:
access control for information security is delegated to the host
operating system and web server. In practice, operating
systems rely either on coarse permission models,
which tend to be ineffective, or on detailed access control
lists, which tend to be cumbersome. This means that information security
in RCloud is a delicate, unsolved matter. More
sophisticated approaches, such as information flow analysis,
and query languages that ensure differential privacy, are
still research topics~\cite{Moore:2011:SAF}.
% We are hopeful that having a richer
% environment for collaboration and information sharing will encourage
% new approaches to information security in visual analytics.
%\gordon{This sounds like we think we are helping by pushing the
%system to the breaking point, is that the idea?}

\paragraph*{Nonforgetful, effortless exploration is hard}
Users reported that RCloud's combined notebook and command line
interface interferes with their familiar use of R for exploration.  In
current R practice, the command prompt is used to try things out.  It
is very fast and there is often no commitment: nothing gets saved by
default, but clean up is automatic.  In contrast, in RCloud's
notebook interface, cells generated doing exploration are saved in the
notebook.  We arrived here from an insistence on reproducibility: it is a
core principle of RCloud that no command should be run without being
saved. In our experience some history steps are only found to be
important in hindsight, and deleting history by default makes it too
easy to lose those steps by accident.

Surprisingly, we found that when users realize that \emph{scratch work
  is getting saved}, the extra versions become a mental burden for
them. Clearly, RCloud has now moved too far to the other side, and the
right solution is somewhere in the middle. An intriguing possibility,
suggested by one of our interviewees, is to implement auditing of data
analyses as it existed in S~\cite{Becker:1988:Auditing}, tying the
data artifacts to the processes that generated them. As a result, the
bookkeeping is reduced to selecting the \emph{data} to be kept, and
the code which generates it can be derived and saved transparently.

%% . This is why we did not implement a simple command line
%% as many users request. We could imagine making an exception for commands
%% that have no side-effects, but in the general case this is no easier than
%% auditing because R is not a purely functional language.

%% The interface is also perceived to be slower, both for computer and for
%% human. First, there is higher latency when commands run, because
%% RCloud saves a cell to a repository before executing it. It
%% may be possible to safely write to the repository behind execution of the
%% same code. Second, and perhaps more detrimental, the interface requires
%% switching between the keyboard and mouse, which can be cumbersome.
%% %\gordon{we don't need to back that up, right? i hate using passive voice.}
%% Shortcut keys can defined as a workaround, but it would be better
%% to reduce the need for them.

%% Other complaints mostly focus on layout, and can probably be addressed by
%% offering better customization, and designs such as tabbed views.

\paragraph*{Discovery is hard} 
With hundreds of users and thousands of notebooks, the notebook tree
becomes unwieldy. The search function does not protect searchers from
obsolete and erroneous notebooks, which by default are retained
forever.  In addition, some important metadata (has this notebook been
recently run, or edited?) is also not easily searchable. In other words,
\emph{textual search} is easy, but \emph{relevance search} is what we
need. Allowing users to \emph{tag} notebooks is a natural idea, but it
relies on users making a conscious effort, which goes somewhat against
our ideal of not asking the user to do work for the system. Rather, we
believe a better alternative is to have the system work for the user,
relying on passive metrics, such as how often a notebook has been run.
More generally, infrastructure for automatic \emph{curation} of the
software artifacts generated by teams seems to be an important avenue
for future work.
